---
title: "5300DTCWriteup"
authors: "Wanchi Peng, Adante Henderson, Andrew Zinkan, Jonathan Burns, Brad Stirling"
output: html_document
date: "2022-08-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

This section loads all required libraries

```{r}
library(fixest)
library(tidyverse)
library(vtable)
library(lubridate)
library(ipumsr)
library(rdrobust)
library(dplyr)
library(ggplot2)
library(margins)
```

## Data Preperation and Wrangling

We pull all of our data from the IPUMS release of the Current Population Survey. We are interested in tracking employment impacts of the Covid Pandemic and therefore have pulled employment data from 2019 through 2021.

```{r}

ddi <- read_ipums_ddi("../data/cps_00006.xml")
emp_data <- read_ipums_micro(ddi)

# Filtering Dataset to Employed Individuals between 2019 and 2021 and creating an "Employment" flag to track employment status. Also filtering out military as they are not typically considered private sector "employed"
emp_data <- emp_data %>% 
  drop_na(IND1990) %>%
  filter(YEAR>=2019 & YEAR <= 2021 & IND1990 > 000 & IND1990 < 940) %>%
  mutate(Employment = case_when(
    EMPSTAT == 10 | EMPSTAT == 12 ~ 1,
    TRUE ~ 0
  )) %>%
  
#Adding date column based off year and month of data and then using date field to calculate the week delta from the start of Covid, (all months after March 2020), utilizing this field to create a flag to track whether a value is Pre-Covid or Post-Covid.
  mutate(date = ymd(paste(YEAR, MONTH, "01"))) %>%
  mutate(weeks = as.numeric(difftime(date,ISOdate(2020,3,1), units="weeks"), units="weeks")) %>%
  filter(MONTH != 3) %>%
  mutate(post_covid = case_when(
    weeks < 0 ~ 0,
    TRUE ~ 1
  )) %>%
  # https://cps.ipums.org/cps/codes/ind_2014_codes.shtml
  mutate(Industry = as.factor(case_when(
    IND1990 > 000 & IND1990 <= 032 ~ "AGRICULTURE, FORESTRY, AND FISHERIES",
    IND1990 >= 040 & IND1990 <= 050 ~ "MINING",
    IND1990 == 060 ~ "CONSTRUCTION",
    IND1990 >= 100 & IND1990 <= 392 ~ "MANUFACTURING",
    IND1990 >= 400 & IND1990 <= 472 ~ "TRANSPORTATION, COMMUNICATIONS, AND OTHER PUBLIC UTILITIES",
    IND1990 >= 500 & IND1990 <= 571 ~ "WHOLESALE TRADE",
    IND1990 >= 580 & IND1990 <= 691 ~ "RETAIL TRADE",
    IND1990 >= 700 & IND1990 <= 712 ~ "FINANCE, INSURANCE, AND REAL ESTATE",
    IND1990 >= 721 & IND1990 <= 760 ~ "BUSINESS AND REPAIR SERVICES",
    IND1990 >= 761 & IND1990 <= 791 ~ "PERSONAL SERVICES",
    IND1990 >= 800 & IND1990 <= 810 ~ "ENTERTAINMENT AND RECREATION SERVICES",
    IND1990 >= 812 & IND1990 <= 893 ~ "PROFESSIONAL AND RELATED SERVICES",
    IND1990 >= 900 & IND1990 <= 932 ~ "PUBLIC ADMINISTRATION",
    TRUE ~ "NIU"
  ))) %>%
  filter(Industry != "NIU") %>%
  mutate(Retail = Industry == "RETAIL TRADE")

# Isolating retail
data_retail <- emp_data %>%
  filter(Retail == 1)

# Isolating non-retail
data_non_retail <- emp_data %>%
  filter(Retail != 1)

# Summarizing by retail[T/F] by date
data_ret_summ <- emp_data %>%
  group_by(date, weeks, Retail) %>%
  summarise(Employment=mean(Employment))

```

## Question One - The impact of Covid on Retail Employment

The first part of the analysis utilized employment data to explore the effect of Covid on the health of the retail industry. The sample consisted of individuals in the retail industry aged 25 to 65 who were classified as part of the labor force (employed or unemployed) during the span of April 2019 to March 2021. Individuals were assigned either a value of 1 or 0 depending on their employment status for each month. These values were then aggregated at the monthly grain to calculate total employment. As the Covid lockdown was staggered throughout March 2021, the month was removed from the analysis. A binary variable for the lockdown months (April 2020 to March 2021) was created as the treatment variable for the model.

```{r}
#We roll up employment to the month level for analysis
q1_data <- data_retail %>% group_by(date, post_covid) %>%
  summarize(monthly_employment = sum(Employment))

sumtable(q1_data)

q1_data %>% ggplot(aes(date, monthly_employment)) +
  geom_line(color = "dark blue") +
  ggtitle("Retail Employment: April 2019 to February 2021") +
  geom_vline(xintercept = as.Date("2020-02-01") - 0.5,
             col = "orange",
             linetype = "dashed"
  ) +
  xlab("Date (excludes March 2020)") +
  ylab("Monthly Employment") +

  annotate(geom = "text",
           x = as.Date("2020-02-01") -50,
           y = 3400,
           label = "Lockdown",
           color = "orange")
```

```{r}
q1_data_month <- q1_data %>%
  filter(month(date) > 2) %>%
  mutate(month = factor(month(date),levels = c(4:12)),
         year = factor(year(date)),
         post_covid = case_when(post_covid == 1 ~ "Post",
                               TRUE ~ "Pre"))
```

```{r}
q1_data_month %>% ggplot(aes(x = month, 
                             y = monthly_employment,
                             col = post_covid)) +
  geom_point() + 
  ggtitle("Monthly Employment Trends: April to December - 2019/2020") +
  labs(col = "Lockdown")
```

```{r}

q1_m1 <-  q1_data %>% feols(monthly_employment ~ post_covid)


q1_m1 <-  q1_data %>% feols(monthly_employment ~ covid_era)

etable(q1_m1)
```

The retail industry saw a drop in monthly employment of 1,298 people post-lockdown, a decrease of 25.8%. This was statistically significant at an alpha of 0.001. Thus, based on the model, there is strong evidence that the lockdown had a negative impact on the health of the retail industry. 

This conclusion is contingent on the assumption that the drop was due to the lockdown and not other factors. While retail employment was trending downwards in the months leading up to the lockdown, this assumption is backed up in the "Retail Employment" graph which shows a steep decline in employment following the lockdown measures in March 2020. Additionally, employment then proceeds to climb upwards as stores reopened in the months following. The "Monthly Employment Trends" graph further supports this assumption, as after lockdown restrictions eased up throughout the summer of 2020, September to December have similar trends compared to the same span in the previous year.

##Question Two - Covid Impact on Retail vs Other Organizations

The second research question is investigating if retail employment differed from other
industries. Several data transformations were generated to compare:

  1. Industry groupings based on the IND1990 categories found here:
      https://cps.ipums.org/cps-action/variables/IND1990#codes_section
  2. Filtering out Not In Universe and Military jobs to focus on the private sector 
  3. A binary variable that is TRUE when the industry is retail, and FALSE otherwise
  4. Aggregated monthly employment rates by industry
  5. Since the lock down began mid-March, all data from March 2020 is filtered
      out to prevent issues due to reporting inconsistencies

For reference, the average employment rates by industry are as follows:

```{r}

# Aggregating monthly data and ensuring
Q2_data_mo <- emp_data %>%
  filter(Industry != "NIU" & Industry != "ACTIVE DUTY MILITARY") %>%
  group_by(Industry, date, Retail, weeks, post_covid) %>%
  summarise(Employment = mean(Employment))

Q2_data_mo %>%
  group_by(Industry) %>%
  summarize("Employment Rate" = mean(Employment))
```

## Plots

A graph of each industry's employment rate over time shows differences across
industries:

```{r}
ggplot(data = Q2_data_mo, aes(x=date, y=Employment, colour=Industry, group=Industry)) +
  geom_point() + geom_line()
```

While retail did not exhibit the strongest declines, it appears to have declined
drastically more than the strongest and indeed was the third worst according to this graph.

Isolating Retail vs all other industries helps visualize this difference:

```{r}
ggplot(data=data_ret_summ, aes(x=date, y=Employment, colour=Retail)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept=ISOdate(2020,3,1))
```
While pre-pandemic employment is a bit lower for retail than other industries, the
overall trend looks very similar. However, the decline immediately after lock downs
was greater both absolutely and proportionately.

Next, rdplots were generated at first and second orders, for all industries and
retail alone.

```{r}
rdplot(Q2_data_mo$Employment, Q2_data_mo$weeks, c=0, p=1, h=24, kernel = "uniform",
       title = "Linear - All Industries", x.label = "Weeks Since March")

rdplot(Q2_data_mo$Employment, Q2_data_mo$weeks, c=0, p=2, h=25, kernel = "uniform",
       title = "Second order - All Industries", x.label = "Weeks Since March")

rdplot(data_retail$Employment, data_retail$weeks, c=0, p=2, h=25, kernel = "uniform",
       title = "Second Order - Retail Only", x.label = "Weeks Since March")
```

## Models

Several models were created to analyze the pandemic, starting with a fixed effect
ols model regressing employment by date and pre/post pandemic, with fixed effects
applied by Industry:

```{r}
model_1 <- feols(Employment ~ weeks * post_covid | Industry, data=Q2_data_mo, vcov="hetero")
etable(model_1)
wald(model_1)
```
The Wald test shows statistically significant coefficients by industry, but we can
investigate further to provide more context via different methods.

Noting that weeks was not individually statistically significant to the model,
four more models were generated using both rdrobust and feols to compare different
structures:

```{r}
model_2 <- feols(Employment ~ weeks * post_covid | Retail, data=Q2_data_mo, se="hetero")
model_2_rd_r <- rdrobust(data_retail$Employment, data_retail$weeks, c=0, p=2, h=25, kernel = "uniform")
model_2_rd_nr <- rdrobust(data_non_retail$Employment, data_non_retail$weeks, c=0, p=2, h=25, kernel = "uniform")
model_3 <- feols(Employment ~ post_covid * Industry, data = Q2_data_mo, se="hetero")
```

Comparing the three feols() models:

```{r}
etable(model_1, model_2)
etable(model_3)
```

These three models show mostly statistically significant coefficients, and a respectable
R2 when regressing Employment on post_covid (i.e. has the pandemic happened) and industry
in model_3.

The rdrobust models allow for more targeted analysis of second order models, controlling
for retail:

```{r}
summary(model_2_rd_r, vcov="hetero")
summary(model_2_rd_nr, vcov="hetero")
```
The coefficient for the retail model shows a 24% decrease in employment rate, and
that of the non-retail model shows a 10% decrease. Comparing these estimates and their
robust 95% confidence intervals, we see that:

#### Retail employment decreased an estimated 14% more than the private sector
#### as a whole, which is statistically significant.

Finally, individual models were created for each industry to demonstrate individual
comparisons. Additional work could be conducted to precisely identify the significance
of differences between pairs of industries, but the conclusion above was determined
to sufficiently answer the research question at hand.

```{r}
IndList <- split(Q2_data_mo, f=Q2_data_mo$Industry)
model_list <- list()
rdrobust_list <- list()

for (i in 1:length(IndList)) {
  indus <- levels(IndList[[i]]$Industry)[i]
  modelname <- paste0(indus, "_model") %>%
    replace(" ", "_")
  
  model_list[[paste0(indus)]] <- assign(modelname,feols(Employment ~ weeks * post_covid, data=IndList[[i]]))
  rdrobust_list[[paste0(indus)]] <- assign(modelname,rdrobust(IndList[[i]]$Employment,IndList[[i]]$weeks,c=0,p=2,h=25,kernel="uniform"))
}

etable(model_list)
for (i in 1:length(rdrobust_list)) {
  print(names(rdrobust_list)[i])
  summary(rdrobust_list[[i]])
}
```

Looking at the etable output, we see that only Personal Services and Entertainment
industries exhibited greater declines in employment than Retail. As noted above,
additional analysis should be conducted if the magnitude and significant of these
specific differences is of interest.

##Question Three - Changes in employment and consumer spending post-lockdown

```{r}
load("../code/Q3_Data_Wrangling.RData")
sumtable(df)
```


First, a final variable, `is_employed`, was added. which classifies EMPSTAT in (10, 12) as `employed` and EMPSTAT in (21,22) as `Unemployed`, and changed pre post to a bool value. Covid status was also changed to a `bool` value. 
```{r}
df <- df %>% mutate(is_employed = case_when(EMPSTAT == 10 ~ '1', 
                                            EMPSTAT == 12 ~ '1',
                                            TRUE ~ '0')) %>% 
  mutate(is_employed = as.numeric(is_employed))
df$is_post <- (df$COVIDSTATUS == 'post-covid')
```


## High Level Frequency stats
Next, frequency stats were added to view the dispersion of our binary dependent variable. As per the table, majority class across the board is employed, and thus there was no large differences in employment among demographic traits, martial status, or industry that was clearly visible.

Because the following factor variables will be used as dummy variables down stream, other variables were removed from the analysis (i.e. to avoid making a dummy variable where a single person is employed in a niche market as this adds more features with little importance). 
```{r}
ggplot(df, aes(x= RACE, color = factor(is_employed))) + geom_bar() + 
  ggtitle("Count of Employment per Race") +
  xlab("Race")
```

$Note:$ Looking at break out by race, there are some counts that could add noise to the data and create many dummy variables. Thus, the sample was adjusted to  the top 10 `RACE` demographics in terms of employment population. 

$Note:$ Filtering data for just our top ten most populated RACE categories.  

```{r, echo=TRUE}
df %>% 
  group_by(RACE) %>% 
  summarise(rec_count = n()) %>% 
  arrange(., desc(rec_count))
df <- df %>% 
  filter(RACE == c(100,200,300,651,802,652,801,803,805,804))
```
Looking at employment by industry
```{r}

ggplot(df, aes(x= indname, color = factor(is_employed))) + geom_bar() + 
  ggtitle("Count of Employment per Industry") +
  theme(axis.text.x = element_text(angle = 90
                                   )) + 
  ggtitle("Count of Employment per Industry") +
  xlab("Industry")
```

$Note:$: As Construction sector had very few employees, it was removed from further analysis to reduce noise

```{r}
ggplot(df, aes(x= Age_broad, color = factor(is_employed))) + geom_bar() +
  ggtitle("Employment per Age Group") +
  xlab("Age Group")
```

$Note:$: Age shows no large discrepancies, with employment. 
```{r}
ggplot(df, aes(x= SEX, color = factor(is_employed))) + geom_bar() +
  ggtitle("Employment per Gender") +
  xlab("Gender")
```

$Note:$: Sex shows no large discrepancies with employment.

```{r}
ggplot(df, aes(x= MARST, color = factor(is_employed))) + geom_bar() +
  ggtitle("Employment per Marital Status") +
  xlab("Marital Status")
```

$Note:$: Marital status shows no large discrepancies with employment.

***
While the above analysis looks at pre and post Covid together the next part examines employment changes over time. 

### _Drop in total employment_

```{r}
agg <- df %>% 
  group_by(yearmo) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>% 
  arrange(., yearmo)
ggplot(data=agg, aes(y = employed_count, x = yearmo))+
  geom_line(alpha = 0.6)+
  geom_vline(xintercept = 202003, colour="#BB0000")+
  labs(y = 'Count of Employeed', x = 'Time')
```

$Note:$ The large fall right after the start of Covid was followed by a large uptick a few months later.
***

Do to the volatility and scale I also want to look at the employment variation as % changes to capture a more normalized metric. 
### _By percentage change to normalize the change_
```{r, warning=FALSE}
agg <- df %>% 
  group_by(yearmo) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>% 
  arrange(., yearmo)
agg <- agg %>% 
  mutate(lag1_value = lag(employed_count, n=1, order_by=yearmo)) %>% 
  mutate(prc_change_emp = (employed_count-lag1_value)/ lag1_value)
ggplot(data=agg, aes(y = prc_change_emp , x = yearmo))+
  geom_line(alpha = 0.6)+
  geom_vline(xintercept = 202003, colour="#BB0000")+
  labs(y = 'Count of Employeed', x = 'Time')
```

$Note:$ There was a  ~30% fall in total employment count following Covid, then a ~15% rebound immediately after. There is no question that employment took a hit after Covid, but the real question is how did it impact different demographics?

***

# Employment before and after covid
In this section, employment numbers and the % change in employment fluctuations after Covid are explored.

### _Drop by Race_

```{r, message=FALSE}
agg <- df %>% 
  group_by(yearmo, RACE) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>% 
  filter(yearmo >= 201950) %>% 
  filter(yearmo <= 202020) %>%
  arrange(.,RACE, yearmo)
ggplot(data=agg, aes(y = log(employed_count), x = yearmo))+
  geom_line(alpha = 0.6, aes(color = factor(RACE)))+
  geom_vline(xintercept = 202003)+
  labs(y = 'Count of Employeed', x = 'Time', color = 'RACE')
```

$Note:$ Most races show a similar level of employment drop, however some races in terms of over all count are more or less employed to begin with, so a similar drop in employment numbers across race might not really be proportional impacts. 

***

To examine this in more detail the % change overtime by race is further explored. 

### _% Drop by RACE_
```{r, echo= FALSE, message= FALSE, warning=FALSE}
agg <- df %>% 
  group_by(yearmo, RACE) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>% 
  # filter(RACE == c(100,200,300,651,652)) %>% 
  filter(yearmo >= 201950) %>% 
  filter(yearmo <= 202020) %>%
  arrange(., yearmo)
  
agg <- agg %>% 
  group_by(RACE) %>% 
  mutate(lag1_value = lag(employed_count, n=1, order_by= yearmo)) %>% 
  mutate(prc_change_emp = (employed_count-lag1_value)/ lag1_value) %>% 
  arrange(., RACE, yearmo)
ggplot(data=agg, aes(y = prc_change_emp, x = yearmo))+
  geom_line(alpha = 0.6, aes(color = factor(RACE)))+
  geom_vline(xintercept = 202003, colour="#BB0000")+
  labs(y = 'Count of Employeed', x = 'Time', color = 'RACE')
```

$Note:$ While some races stay relatively steady and correlated to each other, others show high fluctuations. _Some of these are also races with small sample size so a 1 person change in employment makes a large delta in this graph_.


### _Drop by indname_
```{r, message=FALSE}
agg <- df %>% 
  group_by(yearmo, indname) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>% 
  filter(yearmo >= 201950) %>% 
  filter(yearmo <= 202020) %>%
  arrange(.,indname, yearmo)
ggplot(data=agg, aes(y = employed_count, x = yearmo))+
  geom_line(alpha = 0.6, aes(color = factor(indname)))+
  geom_vline(xintercept = 202003)+
  labs(y = 'Count of Employeed', x = 'Time', color = 'Industry Name')
```

$Note:$ That NA takes a majority representation in our data set and constructions takes a very small sample set. This industry type field looks to have quality errors, due to the large amount of N/A values. (majority of this industry data is actually unknown so we may not use this in modeling)

### _% Drop by indname_
Taking a look by industry type
```{r, message=FALSE}
agg <- df %>% 
  group_by(yearmo, indname) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>% 
  filter(yearmo >= 201950) %>% 
  filter(yearmo <= 202020) %>%
  arrange(., yearmo)
agg <- agg %>% 
  group_by(indname) %>% 
  mutate(lag1_value = lag(employed_count, n=1, order_by= yearmo)) %>% 
  mutate(prc_change_emp = (employed_count-lag1_value)/ lag1_value) %>% 
  arrange(., indname, yearmo)
ggplot(data=agg, aes(y = prc_change_emp, x = yearmo))+
  geom_line(alpha = 0.6, aes(color = factor(indname)))+
  geom_vline(xintercept = 202003, colour="#BB0000")+
  labs(y = 'Count of Employeed', x = 'Time', color = 'indname')
```

$Note:$ This graph shows a more normalized representation of each industry. Assuming that `NA` is a catch all for anything that's not `Agriculture, Forestry, etc`, or  `Construction`, or `Transportation` NA would serve as the market indicator representing all other jobs aside from the ones listed here. If we use industry `NA` as the market norm across industries we can see that Transportation and Agriculture follow the overall market employment trend pretty closely, however constructions does not. This might indicate that Transportation and Agriculture do not systematically differ from that of other job types in terms of the covid impact on employment. 

_Construction though also has a very small sample size, so a small change in the number employed can make the % change look more sever than it is (which is the case here_

### _Drop by age group_
```{r, message=FALSE}
agg <- df %>% 
  group_by(yearmo,Age_broad) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>% 
  filter(yearmo >= 201950) %>% 
  filter(yearmo <= 202020) %>%
  arrange(.,Age_broad, yearmo)
ggplot(data=agg, aes(y = employed_count, x = yearmo))+
  geom_line(alpha = 0.6, aes(color = factor(Age_broad)))+
  geom_vline(xintercept = 202003)+
  labs(y = 'Count of Employeed', x = 'Time', color = 'Age_broad')
```
### _% Drop by age group_
```{r, message=FALSE}
agg <- df %>% 
  group_by(yearmo, Age_broad) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>%
  filter(yearmo >= 201950) %>% 
  filter(yearmo <= 202020) %>%
  arrange(., yearmo)
agg <- agg %>% 
  group_by(Age_broad) %>% 
  mutate(lag1_value = lag(employed_count, n=1, order_by= yearmo)) %>% 
  mutate(prc_change_emp = (employed_count-lag1_value)/ lag1_value) %>% 
  arrange(., Age_broad, yearmo)
ggplot(data=agg, aes(y = prc_change_emp, x = yearmo))+
  geom_line(alpha = 0.6, aes(color = factor(Age_broad)))+
  geom_vline(xintercept = 202003, colour="#BB0000")+
  labs(y = 'Count of Employeed', x = 'Time')
```

$Note:$ After looking at the age groups graphs by count and %  we can see that the change across age groups seems to be highly correlated to each other, indicating that one age group does not behave systematically different than that of another.  

### _Drop by sex_
```{r, message=FALSE}
agg <- df %>% 
  group_by(yearmo,SEX) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>% 
  filter(yearmo >= 201950) %>% 
  filter(yearmo <= 202020) %>%
  arrange(.,SEX, yearmo)
ggplot(data=agg, aes(y = employed_count, x = yearmo))+
  geom_line(alpha = 0.6, aes(color = factor(SEX)))+
  geom_vline(xintercept = 202003)+
  labs(y = 'Count of Employeed', x = 'Time', color = 'SEX')
```  
# _% Drop by sex_
```{r, message=FALSE}
agg <- df %>% 
  group_by(yearmo, SEX) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>%
  filter(yearmo >= 201950) %>% 
  filter(yearmo <= 202020) %>%
  arrange(., yearmo)
agg <- agg %>% 
  group_by(SEX) %>% 
  mutate(lag1_value = lag(employed_count, n=1, order_by= yearmo)) %>% 
  mutate(prc_change_emp = (employed_count-lag1_value)/ lag1_value) %>% 
  arrange(., SEX, yearmo)
ggplot(data=agg, aes(y = prc_change_emp, x = yearmo))+
  geom_line(alpha = 0.6, aes(color = factor(SEX)))+
  geom_vline(xintercept = 202003, colour="#BB0000")+
  labs(y = 'Count of Employeed', x = 'Time')
```

$Note:$ Employment change across sex is also correlated to each other indicating that one group does not behave systematically different than that of another. 

### _ Drop by marital status_
```{r, message=FALSE}
agg <- df %>% 
  group_by(yearmo, MARST) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>% 
  filter(yearmo >= 201950) %>% 
  filter(yearmo <= 202020) %>%
  arrange(.,MARST, yearmo)
ggplot(data=agg, aes(y = employed_count, x = yearmo))+
  geom_line(alpha = 0.6, aes(color = factor(MARST)))+
  geom_vline(xintercept = 202003)+
  labs(y = 'Count of Employeed', x = 'Time', color = 'MARST')
```
### _% Drop by marital status_
```{r, message=FALSE}
agg <- df %>% 
  group_by(yearmo, MARST) %>% 
  summarise(employed_count = sum(as.numeric(is_employed))) %>%
  filter(yearmo >= 201950) %>% 
  filter(yearmo <= 202020) %>%
  arrange(., yearmo)
agg <- agg %>% 
  group_by(MARST) %>% 
  mutate(lag1_value = lag(employed_count, n=1, order_by= yearmo)) %>% 
  mutate(prc_change_emp = (employed_count-lag1_value)/ lag1_value) %>% 
  arrange(., MARST, yearmo)
ggplot(data=agg, aes(y = prc_change_emp, x = yearmo))+
  geom_line(alpha = 0.6, aes(color = factor(MARST)))+
  geom_vline(xintercept = 202003, colour="#BB0000")+
  labs(y = 'Count of Employeed', x = 'Time')
```

$Note:$ Employment change across marital is also correlated to each other indicating that one group does not behave systematically different than that of another.

Based on the SEX, AGE graphical relationship we can see that they move in sync, marital status however has more variation in select statuses. Due to these findings Age and sex variables were dropped from the further analyses.  

# Modeling 

### _binary effects (no interaction term)_
The below models show simple linear models with factor variables. This model does _NOT_ look at pre and post Covid differences, it is simply looking at the factor's coefficients on employment as it applies to both pre and post Covid.
```{r}
lpm1 <- feols(is_employed ~ factor(RACE) + factor(SEX) + factor(Age_broad) + factor(MARST) + factor(Education), 
              data = df, se='hetero')
etable(lpm1, digits = 3)
```
_Interpretation_:

- RACE Categories Coefs have a negative effect on employment regardless of if it was pre or post Covid, relative to the `RACE 100` category which is "white". Implying the top 10 races are predicted to be employed less than that of the race "white". Many of these coefficients are statistically significant. 

- Marital status classes are all statistically significant and have a negative coefficient, relative to that of the `Marital status 1` which is "Married, spouse present". This implied that non married statuses are predicted to be employed less than that of the married status. 
- Age groups have little to no statistical significance across age groups and same with sex. This implied that age does not really play a significant factor in employment status.  


## Predicted values
### _RACE_
```{r}
df <- df %>% 
  mutate(lpm_fitted = predict(lpm1))
ggplot(data = df, mapping = aes(x=RACE, y=lpm_fitted))+
  geom_point()+
  geom_smooth(color = 'red', se=FALSE)+
  theme_bw()
```

$Note:$ This graph shows the predicted value for each race category for both pre and post covid. Notice that certain races are have a lower predicted likelihood for employment status.

### _Education_
```{r}
ggplot(data = df, mapping = aes(x=Education, y=lpm_fitted))+
  geom_point()+
  geom_smooth(color = 'red', se=FALSE)+
  theme_bw()
```

$Note:$ A large difference in doctorate level but not a huge difference between high school and undergrad in terms of predicted employment status.

### _MARST_
```{r}
ggplot(data = df, mapping = aes(x=MARST, y=lpm_fitted))+
  geom_point()+
  geom_smooth(color = 'red', se=FALSE)+
  theme_bw()
```

$Note:$ Fairly constant predictive value across marital status but notice that "1" (married spouse present) have the highest predicted value and "3" (separated) and "6" (single) have the lowest predicted probability. 

***


### Modeling with interaction term
Creates dummy education variable
```{r}
lpm3 <- feols(is_employed ~ is_post + Education, data = df, se='hetero')
lpm4 <- feols(is_employed ~  is_post + Education + is_post:Education, data = df, se='hetero')
etable(lpm3,lpm4)
```

_Interpretation lpm3_

We have an intercept for our index variable at 0.9904 which is statistically significant.

- is_postTRUE = -0.0438 indicating that the over all effect of covid had a -4.38 percentage point impact on employment.

- doctor = 0.0254 indicating that the over all effect of having a doctorate has a 2.54 percentage point impact on employment relative to bachelors degree.

- highschool = -0.0161 indicating that the over all effect of having a highschool education or less has a -1.61 percentage point impact on employment relative to bachelors degree. 

$Note:$ All of which are statistically significant at the 0.01% level

_Interpretation lpm4_

This model looks at the interaction term between education and is_post covid. 

- The effect of being a doctor vs having a bachelors post covid is  0.0167 indicating that there is a positive 4.5 percentage point increase associated with being a doctor post covid relative to a bachelors.(this is statistically significant at the 0.01% level). This same effect pre-Covid was 0.0045 or 0.45% percentage points. This results in a Covid effect of 0.04 increase in employment. 

- The effect of having a highschool degree post Covid is -0.0060 indicating that there is a -0.6 percentage point effect associated with highschool degree post Covid relative to a bachelors (this is statistically significant at the 0.01% level). This same effect pre-covid was -0.0135 or -1.35% percentage points relative to a bachelors degree. This results in a covid effect of 0.75 percentage points for high school degrees on employment.


### _RACE_
```{r}
lpm10 <- feols(is_employed ~ is_post + RACE + is_post:RACE, data = df, se='hetero')
etable(lpm10)
```


_Interpretation lpm10_

This is an identical interpretation as above but for the `RACE` variables relative to RACE "White". 

- The effect of RACE200 (black) vs RACE100 "White" post Covid is -0.0154 indicating that there is a negative -1.54 percentage point decrease associated with being black post Covid, relative to being white post Covid on employment rates (this is statistically significant at the 0.1% level).

- Additionally Race200 (black pre-Covid) had a -2.13 percentage point difference compared to Race 100 (white pre-Covid).

- The effect of RACE300 (American Indian) vs RACE100 "White" post covid is -0.0345 indicating that there is a negative -3.45 percentage point decrease associated with being American Indian post Covid, relative to being white post Covid. (this is statistically significant at the 0.1% level). 

- Additionally Race300 (American Indian pre-covid) had a -4.19 percentage point difference compared to Race 100 (white pre-covid).


$Note:$ Covid has a stronger effect on these races than white, also not that native american impact was stronger that the impact to the black population by ~ 2x.



### _Matital Status_
```{r}
lpm11 <- feols(is_employed ~ is_post + MARST, data = df, se='hetero')
lpm12 <- feols(is_employed ~ MARST + is_post + is_post:MARST, data = df, se='hetero')
etable(lpm11,lpm12)
```

_Interpretation lmp12_

This model looks at the interaction term between marital status and covid. These results are all relative to MARST = 1 or married with spouse present. 

- The effect of having a marital status = 3 (separated) vs status = 1 (married) post covid is -0.0517 indicating that there is a -5.17 percentage point change associated with being separated post Covid (this is statistically significant at the 1% level).

- The coefficient on being separated pre covid is -0.0524 or -5.24 percentage point. 

The same observations occurs for MARST 6 (single) as well resulting in a covid effect of -3.14 percentage points.

# Summary

Based on our analysis of demographics and the effect of covid on each demographic we can conclude there there are select demographic traits that are more impacted than others by Covid.

Out of the select demographic variables used for analysis (RACE, Education, and Marital Status) we can conclude that select RACE groups, Education status, and marital statuses were more heavily impact by Covid. 

Education:

- Doctoral degree was impacted by Covid by 4.5 percentage points.
- High school degree or less was impacted by COVID by -0.6 percentage points

Race:

When compared to RACE100 (white)

- The effect of Covid on a Black individual was -1.54 percentage points relative to white. 
- The effect of covid on a Native American individual was -3.45 percentage points relative to white.

Marital Status:

When compared to being married.

- The effect of Covid on a separated individual was -2.12 percentage points relative to married. 
- The effect of Covid on a single individual was -3.14 percentage points relative to married. 

Using employment as a pseudo measure for spending power, we would recommend that a retail organization (post Covid) focus their retail and marketing strategy on a select demographic with employment that was less impacted by the effect of Covid. This assumes that the retail organization is not focused on specific target audience or demographic, and that the retail organization in mind prefers quantity of customers, and is profit seeking. In a changing post Covid market environment, we would recommend a retail strategy that's focused on individuals with the following traits: well educated, white, and married. 

