---
title: "5300DTCWriteup"
output: html_document
date: "2022-08-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

This section loads all required libraries

```{Libraries}
library(fixest)
library(tidyverse)
library(vtable)
library(lubridate)
library(ipumsr)
library(rdrobust)
library(dplyr)
library(ggplot2)
library(marginaleffects)
```

## Data Preperation and Wrangling

We pull all of our data from the IPUMS release of the Current Population Survey. We are interested in tracking employment impacts of the Covid Pandemic and therefore have pulled employment data from 2019 through 2021

```{Data Preperation}
ddi <- read_ipums_ddi("../data/cps_00004.xml")
emp_data <- read_ipums_micro(ddi)

# Filtering Dataset to Employed Individuals between 2019 and 2021 and creating an "Employment" flag to track employment status. Also filtering out military as they are not typically considered private sector "employed"
emp_data <- emp_data %>% 
  drop_na(IND1990) %>%
  filter(YEAR>=2019 & YEAR <= 2021 & IND1990 > 000 & IND1990 < 940) %>%
  mutate(Employment = case_when(
    EMPSTAT == 10 | EMPSTAT == 12 ~ 1,
    TRUE ~ 0
  )) %>%
  
#Adding date column based off year and month of data and then using date field to calculate the week delta from the start of Covid, (all months after March 2020), utilizing this field to create a flag to track whether a value is Pre-Covid or Post-Covid.
  mutate(date = ISOdate(YEAR,MONTH,1)) %>%
  mutate(weeks = as.numeric(difftime(date,ISOdate(2020,3,1), units="weeks"), units="weeks")) %>%
  filter(MONTH != 3) %>%
  mutate(disc_id = case_when(
    weeks < 0 ~ 0,
    TRUE ~ 1
  )) %>%
  # https://cps.ipums.org/cps/codes/ind_2014_codes.shtml
  mutate(Industry = as.factor(case_when(
    IND1990 > 000 & IND1990 <= 032 ~ "AGRICULTURE, FORESTRY, AND FISHERIES",
    IND1990 >= 040 & IND1990 <= 050 ~ "MINING",
    IND1990 == 060 ~ "CONSTRUCTION",
    IND1990 >= 100 & IND1990 <= 392 ~ "MANUFACTURING",
    IND1990 >= 400 & IND1990 <= 472 ~ "TRANSPORTATION, COMMUNICATIONS, AND OTHER PUBLIC UTILITIES",
    IND1990 >= 500 & IND1990 <= 571 ~ "WHOLESALE TRADE",
    IND1990 >= 580 & IND1990 <= 691 ~ "RETAIL TRADE",
    IND1990 >= 700 & IND1990 <= 712 ~ "FINANCE, INSURANCE, AND REAL ESTATE",
    IND1990 >= 721 & IND1990 <= 760 ~ "BUSINESS AND REPAIR SERVICES",
    IND1990 >= 761 & IND1990 <= 791 ~ "PERSONAL SERVICES",
    IND1990 >= 800 & IND1990 <= 810 ~ "ENTERTAINMENT AND RECREATION SERVICES",
    IND1990 >= 812 & IND1990 <= 893 ~ "PROFESSIONAL AND RELATED SERVICES",
    IND1990 >= 900 & IND1990 <= 932 ~ "PUBLIC ADMINISTRATION",
    TRUE ~ "NIU"
  ))) %>%
  filter(Industry != "NIU") %>%
  mutate(Retail = Industry == "RETAIL TRADE")

# Isolating retail
data_retail <- emp_data %>%
  filter(Retail == 1)

# Isolating non-retail
data_non_retail <- emp_data %>%
  filter(Retail != 1)

# Summarizing by retail[T/F] by date
data_ret_summ <- emp_data %>%
  group_by(date, weeks, Retail) %>%
  summarise(Employment=mean(Employment))

```

## Question One - The impact of Covid on Retail Employment

We investigate the question "How has COVID affected the health of the retail industry, as measured by employment?" below

```{Data Summary}
#We roll up employment to the month level for analysis
q1_data <- data_retail %>% group_by(date, disc_id) %>%
  summarize(monthly_employment = sum(Employment))
```

The first part of the analysis utilized employment data to explore the effect of COVID on the health of the retail industry. The sample consisted of individuals in the retail industry aged 25 to 65 who were classified as part of the labor force (employed or unemployed) during the span of April 2019 to March 2021. Individuals were assigned either a value of 1 or 0 depending on their employment status for each month. These values were then aggregated at the monthly grain to calculate total employment. As the COVID lockdown was staggered throughout March 2021, the month was removed from the analysis. A binary variable for the lockdown months (April 2020 to March 2021) was created as the treatment variable for the model.

```{r}
q1_data %>% ggplot(aes(date, monthly_employment)) +
  geom_line(color = "dark blue") +
  ggtitle("Retail Employment: April 2019 to February 2021") +
  geom_vline(xintercept = as.Date("2020-02-01") - 0.5,
             col = "orange",
             linetype = "dashed"
  ) +
  annotate(geom = "text",
           x = as.Date("2020-02-01") -50,
           y = 3400,
           label = "Lockdown",
           color = "orange") +
  xlab("Date (excludes March 2020)") +
  ylab("Monthly Employment")
```

```{r}
q1_data_month <- q1_data %>%
  filter(month(date) > 2) %>%
  mutate(month = factor(month(date),levels = c(4:12)),
         year = factor(year(date)),
         covid_era = case_when(covid_era == 1 ~ "Post",
                               TRUE ~ "Pre"))
```

```{r}
q1_data_month %>% ggplot(aes(x = month, 
                             y = monthly_employment,
                             col = covid_era)) +
  geom_point() + 
  ggtitle("Monthly Employment Trends: April to December - 2019/2020") +
  labs(col = "Lockdown")
```

```{r}
q1_m1 <-  q1_data %>% feols(monthly_employment ~ covid_era)

etable(q1_m1)
```

The retail industry saw a drop in monthly employment of 1,298 people post lockdown, a decrease of 25.8%. This was statistically significant at an alpha of 0.001. Thus, based on the model, there is strong evidence that the lockdown had a negative impact on the health of the retail industry. 

This conclusion is contingent on the assumption that the drop was due to the lockdown and not other factors. While retail employment was trending downwards in the months leading up to the lockdown, this assumption is backed up in the "Retail Employment" graph which shows a steep decline in employment following the Lockdown measures in March 2020. Additionally, employment then proceeds to climb upwards as stores reopened in the months following. The "Monthly Employment Trends" graph further supports this assumption, as after lockdown restrictions eased up throughout the summer of 2020, September to December have similar trends compared to the same span in the previous year.

##Question Two - Covid Impact on Retail vs Other Organizations

The second research question is investigating if retail employment differed from other
industries. Several data transformations were generated to compare:

  1. Industry groupings based on the IND1990 categories found here:
      https://cps.ipums.org/cps-action/variables/IND1990#codes_section
  2. Filtering out Not In Universe and Military jobs to focus on the private sector 
  3. A binary variable that is TRUE when the industry is retail, and FALSE otherwise
  4. Aggregated monthly employment rates by industry
  5. Since the lock down began mid-March, all data from March 2020 is filtered
      out to prevent issues due to reporting inconsistencies

For reference, the average employment rates by industry are as follows:

```{r code}

# Aggregating monthly data and ensuring
Q2_data_mo <- emp_data %>%
  filter(Industry != "NIU" & Industry != "ACTIVE DUTY MILITARY") %>%
  group_by(Industry, date, Retail, weeks, disc_id) %>%
  summarise(Employment = mean(Employment))

Q2_data_mo %>%
  group_by(Industry) %>%
  summarize("Employment Rate" = mean(Employment))
```

## Plots

A graph of each industry's employment rate over time shows differences across
industries:

```{r}
ggplot(data = Q2_data_mo, aes(x=date, y=Employment, colour=Industry, group=Industry)) +
  geom_point() + geom_line()
```

While retail did not exhibit the strongest declines, it appears to have declined
drastically more than the strongest and indeed was the third worst according to this graph.

Isolating Retail vs all other industries helps visualize this difference:

```{r}
ggplot(data=data_ret_summ, aes(x=date, y=Employment, colour=Retail)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept=ISOdate(2020,3,1))
```
While pre-pandemic employment is a bit lower for retail than other industries, the
overall trend looks very similar. However, the decline immediately after lock downs
was greater both absolutely and proportionately.

Next, rdplots were generated at first and second orders, for all industries and
retail alone.

```{r}
rdplot(Q2_data_mo$Employment, Q2_data_mo$weeks, c=0, p=1, h=24, kernel = "uniform",
       title = "Linear - All Industries", x.label = "Weeks Since March")

rdplot(Q2_data_mo$Employment, Q2_data_mo$weeks, c=0, p=2, h=25, kernel = "uniform",
       title = "Second order - All Industries", x.label = "Weeks Since March")

rdplot(data_retail$Employment, data_retail$weeks, c=0, p=2, h=25, kernel = "uniform",
       title = "Second Order - Retail Only", x.label = "Weeks Since March")
```

## Models

Several models were created to analyze the pandemic, starting with a fixed effect
ols model regressing employment by date and pre/post pandemic, with fixed effects
applied by Industry:

```{r}
model_1 <- feols(Employment ~ weeks * disc_id | Industry, data=Q2_data_mo, vcov="hetero")
etable(model_1)
wald(model_1)
```
The Wald test shows statistically significant coefficients by industry, but we can
investigate further to provide more context via different methods.

Noting that weeks was not individually statistically significant to the model,
four more models were generated using both rdrobust and feols to compare different
structures:

```{r}
model_2 <- feols(Employment ~ weeks * disc_id | Retail, data=Q2_data_mo, se="hetero")
model_2_rd_r <- rdrobust(data_retail$Employment, data_retail$weeks, c=0, p=2, h=25, kernel = "uniform")
model_2_rd_nr <- rdrobust(data_non_retail$Employment, data_non_retail$weeks, c=0, p=2, h=25, kernel = "uniform")
model_3 <- feols(Employment ~ disc_id * Industry, data = Q2_data_mo, se="hetero")
```

Comparing the three feols() models:

```{r}
etable(model_1, model_2)
etable(model_3)
```

These three models show mostly statistically significant coefficients, and a respectable
R2 when regressing Employment on disc_id (i.e. has the pandemic happened) and industry
in model_3.

The rdrobust models allow for more targeted analysis of second order models, controlling
for retail:

```{r}
summary(model_2_rd_r, vcov="hetero")
summary(model_2_rd_nr, vcov="hetero")
```
The coefficient for the retail model shows a 24% decrease in employment rate, and
that of the non-retail model shows a 10% decrease. Comparing these estimates and their
robust 95% confidence intervals, we see that:

#### Retail employment decreased an estimated 14% more than the private sector
#### as a whole, which is statistically significant.

Finally, individual models were created for each industry to demonstrate individual
comparisons. Additional work could be conducted to precisely identify the significance
of differences between pairs of industries, but the conclusion above was determined
to sufficiently answer the research question at hand.

```{r}
IndList <- split(Q2_data_mo, f=Q2_data_mo$Industry)
model_list <- list()
rdrobust_list <- list()

for (i in 1:length(IndList)) {
  indus <- levels(IndList[[i]]$Industry)[i]
  modelname <- paste0(indus, "_model") %>%
    replace(" ", "_")
  
  model_list[[paste0(indus)]] <- assign(modelname,feols(Employment ~ weeks * disc_id, data=IndList[[i]]))
  rdrobust_list[[paste0(indus)]] <- assign(modelname,rdrobust(IndList[[i]]$Employment,IndList[[i]]$weeks,c=0,p=2,h=25,kernel="uniform"))
}

etable(model_list)
for (i in 1:length(rdrobust_list)) {
  print(names(rdrobust_list)[i])
  summary(rdrobust_list[[i]])
}
```

Looking at the etable output, we see that only Personal Services and Entertainment
industries exhibited greater declines in employment than Retail. As noted above,
additional analysis should be conducted if the magnitude and significant of these
specific differences is of interest.