---
title: "Q2_md"
author: "Jonathan Burns"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE,message=FALSE)
library(tidyverse)
library(ipumsr)
library(rdrobust)
library(fixest)
library(vtable)
source("Q2_Data_Wrangling.R", local = knitr::knit_global())
```

## Research question and data summary

The second research question is investigating if retail employment differed from other
industries. Several data transformations were generated to compare:

  1. Industry groupings based on the IND1990 categories found here:
      https://cps.ipums.org/cps-action/variables/IND1990#codes_section
  2. Filtering out Not In Universe and Military jobs to focus on the private sector 
  3. A binary variable that is TRUE when the industry is retail, and FALSE otherwise
  4. Aggregated monthly employment rates by industry
  5. Since the lock down began mid-March, all data from March 2020 is filtered
      out to prevent issues due to reporting inconsistencies

For reference, the average employment rates by industry are as follows:

```{r code}
Q2_data_mo %>%
  group_by(Industry) %>%
  summarize("Employment Rate" = mean(Employment))
```

## Plots

A graph of each industry's employment rate over time shows differences across
industries:

```{r}
ggplot(data = Q2_data_mo, aes(x=date, y=Employment, colour=Industry, group=Industry)) +
  geom_point() + geom_line()
```

While retail did not exhibit the strongest declines, it appears to have declined
drastically more than the strongest and indeed was the third worst according to this graph.

Isolating Retail vs all other industries helps visualize this difference:

```{r}
ggplot(data=data_ret_summ, aes(x=date, y=Employment, colour=Retail)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept=ISOdate(2020,3,1))
```
While pre-pandemic employment is a bit lower for retail than other industries, the
overall trend looks very similar. However, the decline immediately after lock downs
was greater both absolutely and proportionately.

Next, rdplots were generated at first and second orders, for all industries and
retail alone.

```{r}
rdplot(Q2_data_mo$Employment, Q2_data_mo$weeks, c=0, p=1, h=24, kernel = "uniform",
       title = "Linear - All Industries", x.label = "Weeks Since March")

rdplot(Q2_data_mo$Employment, Q2_data_mo$weeks, c=0, p=2, h=25, kernel = "uniform",
       title = "Second order - All Industries", x.label = "Weeks Since March")

rdplot(data_retail$Employment, data_retail$weeks, c=0, p=2, h=25, kernel = "uniform",
       title = "Second Order - Retail Only", x.label = "Weeks Since March")
```

## Models

Several models were created to analyze the pandemic, starting with a fixed effect
ols model regressing employment by date and pre/post pandemic, with fixed effects
applied by Industry:

```{r}
model_1 <- feols(Employment ~ weeks * disc_id | Industry, data=Q2_data_mo, vcov="hetero")
etable(model_1)
wald(model_1)
```
The Wald test shows statistically significant coefficients by industry, but we can
investigate further to provide more context via different methods.

Noting that weeks was not individually statistically significant to the model,
four more models were generated using both rdrobust and feols to compare different
structures:

```{r}
model_2 <- feols(Employment ~ weeks * disc_id | Retail, data=Q2_data_mo, se="hetero")
model_2_rd_r <- rdrobust(data_retail$Employment, data_retail$weeks, c=0, p=2, h=25, kernel = "uniform")
model_2_rd_nr <- rdrobust(data_non_retail$Employment, data_non_retail$weeks, c=0, p=2, h=25, kernel = "uniform")
model_3 <- feols(Employment ~ disc_id * Industry, data = Q2_data_mo, se="hetero")
```

Comparing the three feols() models:

```{r}
etable(model_1, model_2)
etable(model_3)
```

These three models show mostly statistically significant coefficients, and a respectable
R2 when regressing Employment on disc_id (i.e. has the pandemic happened) and industry
in model_3.

The rdrobust models allow for more targeted analysis of second order models, controlling
for retail:

```{r}
summary(model_2_rd_r, vcov="hetero")
summary(model_2_rd_nr, vcov="hetero")
```
The coefficient for the retail model shows a 24% decrease in employment rate, and
that of the non-retail model shows a 10% decrease. Comparing these estimates and their
robust 95% confidence intervals, we see that:

#### Retail employment decreased an estimated 14% more than the private sector
#### as a whole, which is statistically significant.

Finally, individual models were created for each industry to demonstrate individual
comparisons. Additional work could be conducted to precisely identify the significance
of differences between pairs of industries, but the conclusion above was determined
to sufficiently answer the research question at hand.

```{r}
IndList <- split(Q2_data_mo, f=Q2_data_mo$Industry)
model_list <- list()
rdrobust_list <- list()

for (i in 1:length(IndList)) {
  indus <- levels(IndList[[i]]$Industry)[i]
  modelname <- paste0(indus, "_model") %>%
    replace(" ", "_")
  
  model_list[[paste0(indus)]] <- assign(modelname,feols(Employment ~ weeks * disc_id, data=IndList[[i]]))
  rdrobust_list[[paste0(indus)]] <- assign(modelname,rdrobust(IndList[[i]]$Employment,IndList[[i]]$weeks,c=0,p=2,h=25,kernel="uniform"))
}

etable(model_list)
for (i in 1:length(rdrobust_list)) {
  print(names(rdrobust_list)[i])
  summary(rdrobust_list[[i]])
}
```

Looking at the etable output, we see that only Personal Services and Entertainment
industries exhibited greater declines in employment than Retail. As noted above,
additional analysis should be conducted if the magnitude and significant of these
specific differences is of interest.